---
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
  html_document: default
---
# Analysis of the Effects of Vitamin C to Guinea's Pigs Tooth Growth
### *Lou Marvin Caraig*

## Overview

The purpose of this analyis is to evaluate the effects of vitamin C to tooth
growth of Guinea's pig with different combinations of supplement type and
dosage.


## Data exploration and assumptions

```{r tablePlotslibraries, echo=F, results='hide', message=F, warning=F}
library(xtable)
options(xtable.comment=FALSE)
library(ggplot2)
library(gridExtra)
```

```{r invisibleFigrs, echo=F, results='hide', message=F, warning=F}
library(kfigr)

invisible(figr('bySupplementTypePlot', type='Figure'))
invisible(figr('byDoseAmountPlot', type='Figure'))
invisible(figr('byBothPlot', type='Figure'))

invisible(figr('datasetSummary', type='Table'))
invisible(figr('tTestOjGreaterVc', type='Table'))
invisible(figr('tTestBiggerDosesGreaterSmallerDoses', type='Table'))
invisible(figr('cohenTestBiggerDosesGreaterSmallerDoses', type='Table'))
invisible(figr('tTestBoth', type='Table'))

invisible(figr('bySupplementTypePlotCode', type='Code'))
invisible(figr('byDoseAmountPlotCode', type='Code'))
invisible(figr('byBothPlotCode', type='Code'))
invisible(figr('datasetSummaryCode', type='Code'))
invisible(figr('tTestOjGreaterVcCode', type='Code'))
invisible(figr('tTestBiggerDosesGreaterSmallerDosesCode', type='Code'))
invisible(figr('cohenTestBiggerDosesGreaterSmallerDosesCode', type='Code'))
invisible(figr('tTestBothCode', type='Code'))
```

```{r echo=F}
figCaption1 <- paste('On the left the two boxplots.',
                     'On the right the two smoothed density functions.')
figCaption2 <- paste('On the left the three boxplots.',
                     'On the right the three smoothed density functions.')
figCaption3 <- paste('On the left the scatterplot mapping dose amount to tooth length',
                     'with different colors for different supplement type',
                     'with regression line. On the right the couples of boxplots,',
                     'one for each supplement type, for each dose amount.')
```

The `ToothGrowth` dataset can be loaded by simply loading the `datasets`
library, while for the analysis we'll also use `dplyr` and `lsr`:

```{r libraries1, results='hide', message=F, warning=F}
library(datasets)
library(dplyr)
library(lsr)
```

The dataset has `r nrow(ToothGrowth)` rows and `r ncol(ToothGrowth)` columns.
The variable `len` represent the length of odontoblasts (teeth) in micron,
`supp` represent the supplement type of the vitamin C and `dose` represent the
amount of vitamin C dosed in milligrams. The variable `supp` is categorical and
can be equal to `VC` (vitamin C dosed through ascorbic acid) or `OJ` (vitamin C
dosed through orange juice). Here's a summary of the dataset:

```{r datasetSummary, echo=F, results='asis', anchor='Table'}
print(xtable(summary(ToothGrowth), caption='Dataset summary.'))
```

In the following analysis we will perform t-tests and multiple t-tests. For
these tests to be valid we will assume that the observations of the tooth
length are iid drawn from a quasi-normal distribution (roughly symmetric and
mound-shaped), and that the samples are representative for the entire
population. This assumption is made also when the observations are grouped by
supplement type and dose amount.


## Comparison of tooth growth by supplement type, given dose amount

Before defining any hypothesis we first plot the tooth growth by supplement
type. As we can see in `r figr('bySupplementTypePlot', prefix=T)`
(generated with `r figr('bySupplementTypePlotCode', prefix=T)`) dosing
vitamin C through orange juice seems more effective than ascorbic acid. Hence
we test the hypothesis that the difference of means between the tooth length
of those treated with orange juice and those treated with ascorbic acid is
greater than 0. The test is performed at a 0.05 significance level.

```{r bySupplementTypePlot, echo=F, fig.pos='!h', fig.width=7, fig.height=2, fig.align='center', anchor='Figure', fig.cap=figCaption1}
p1 <- ggplot(data=ToothGrowth, mapping=aes(supp, len, colour=supp)) +
    geom_boxplot() +
    geom_jitter(width=0.25) +
    labs(x='Supplement type', y='Tooth length')
p2 <- ggplot(data=ToothGrowth, mapping=aes(x=len, y=..density.., colour=supp)) +
    geom_density() +
    labs(x='Tooth length', y='Density')
grid.arrange(p1, p2, ncol=2, top='Tooth length by supplement type')
```

```{r tTestOjGreaterVc, echo=F, results='asis', anchor='Table'}
oj <- (ToothGrowth %>% filter(supp=='OJ') %>% select(len))$len
vc <- (ToothGrowth %>% filter(supp=='VC') %>% select(len))$len

out <- t.test(oj, vc, alternative='greater')
outDt <- with(out, data.frame(t=statistic, dof=parameter, p=p.value,
                              ci=paste('(', format(conf.int[[1]]), ', ',
                                       format(conf.int[[2]]), ')', sep='')))
colnames(outDt) <- c('t-value', 'degrees of freedom', 'p-value', 'confidence-interval')
print(xtable(outDt, caption=paste('Result of t-test of the comparison of tooth growth',
                                  'by supplement type.')))
```

```{r pValueOjVsVcAux, echo=F}
pValueAux <- outDt$`p-value`
```

As we can see from `r figr('tTestOjGreaterVc', prefix=T)` (calculated with
`r figr('tTestOjGreaterVcCode', prefix=T)`) given that the p-value is equal to
`r pValueAux` we can reject the null hypothesis, meaning that orange juice
affects tooth growth more then ascorbic acid. We can also calculate the effect
size using Cohen's D with `cohensD(oj, vc)`, and it results `r cohensD(oj, vc)`
which can be interpreted as a medium effect.


## Comparison of tooth growth by dose amount, given type of supplement

Now we plot the tooth growth by dose amount. As we can see
in `r figr('byDoseAmountPlot', prefix=T)` (generated with
`r figr('byDoseAmountPlotCode', prefix=T)`) bigger doses of vitamin C seem more
effective than smaller ones. Hence we test the hypothesis that the difference
of means between the tooth length of those treated with bigger doses and those
treated with smaller ones is greater than 0. The test is performed at a 0.05
significance level. Given that we have three groups (three different possible
values of `dose`) we perform a multiple t-test using Bonferroni correction.

```{r byDoseAmountPlot, echo=F, fig.pos='!h', fig.width=7, fig.height=2, fig.align='center', anchor='Figure', fig.cap=figCaption2}
p1 <- ggplot(data=ToothGrowth, mapping=aes(factor(dose), len, colour=factor(dose))) +
    geom_boxplot() +
    geom_jitter(width=0.25) +
    labs(x='Dose amount', y='Tooth length')
p2 <- ggplot(data=ToothGrowth, mapping=aes(x=len, y=..density.., colour=factor(dose))) +
    geom_density() +
    labs(x='Tooth length', y='Density')
grid.arrange(p1, p2, ncol=2, top='Tooth length density by dose amount')
```

```{r tTestBiggerDosesGreaterSmallerDoses, echo=F, results='asis', anchor='Table'}
out <- with(ToothGrowth,
     pairwise.t.test(x=len, g=dose, p.adjust.method='bonferroni', alternative='greater'))
outPValue <- out$p.value
rownames(outPValue) <- paste('dose', rownames(outPValue))
colnames(outPValue) <- paste('dose', colnames(outPValue))
print(xtable(outPValue, display=rep('E', ncol(outPValue) + 1),
             caption=paste('Result of multiple t-test of the comparison of',
                           'tooth growth by dose amount.')))
```

As we can see from `r figr('tTestBiggerDosesGreaterSmallerDoses', prefix=T)`
(calculated with `r figr('tTestBiggerDosesGreaterSmallerDosesCode', prefix=T)`)
all the comparisons (dose 1 vs. dose 0.5, dose 2 vs. dose 0.5,
dose 2 vs dose 1) are statistically significant at a significance level of 0.05
even using such a conservative p-value adjusment method. This means that bigger
doses affects tooth growth more then smaller ones. In
`r figr('cohenTestBiggerDosesGreaterSmallerDoses', prefix=T)` (calculated with
`r figr('cohenTestBiggerDosesGreaterSmallerDosesCode', prefix=T)`) we can see
that the effect size is pretty high in each case especially in the first case
if we consider that the dosage difference is only 0.5 despite the third case
where the dosage difference is 1.0.

```{r cohenTestBiggerDosesGreaterSmallerDoses, echo=F, results='asis', anchor='Table'}
dose0_5 <- (ToothGrowth %>% filter(dose==0.5) %>% select(len))$len
dose1_0 <- (ToothGrowth %>% filter(dose==1.0) %>% select(len))$len
dose2_0 <- (ToothGrowth %>% filter(dose==2.0) %>% select(len))$len

dt <- data.frame(d05_d10=cohensD(dose0_5, dose1_0),
                 d05_d20=cohensD(dose0_5, dose2_0),
                 d10_d20=cohensD(dose1_0, dose2_0))
colnames(dt) <- c(
    'dose 0.5 vs. dose 1.0', 'dose 0.5 vs. dose 2.0', 'dose 1.0 vs. dose 2.0')
rownames(dt) <- ''
print(xtable(dt, caption=paste('Effect size of the difference of tooth growth',
                               'between different dose amount.')))
```


## Comparision of tooth growth by supplement type and dose amount

We can finally consider the tooth growth both by supplement type and dose
amount. As we can see in `r figr('byBothPlot', prefix=T)` (generatedy with
`r figr('byBothPlotCode', prefix=T)`) it seems that in some cases the tooth
length achieved using ascorbic acid with a big dose is the same when using orange
juice with a smaller dose. Hence it's interesting to perform a bidirectional
hypothesis test on the difference of means considering both supplement type and
dose amount. The test is performed at a 0.05 significance level. As before we
have more than two groups so we perform a multiple t-test using Bonferroni
correction. By looking at `r figr('tTestBoth', prefix=T)` (calculated with
`r figr('tTestBothCode', prefix=T)`) we can confirm our supposition. Infact we
can't reject the null hypothesis when doing the following comparisons:
VC-1 vs. OJ-0.5, VC-2 vs. OJ-1, VC-2 OJ-2.


```{r byBothPlot, results='hide', echo=F, warnings=F, fig.pos='!h', fig.width=7, fig.height=2, fig.align='center', anchor='Figure', fig.cap=figCaption3}
p1 <- ggplot(data=ToothGrowth, mapping=aes(dose, len, colour=supp)) +
    geom_point() +
    geom_smooth(method='lm') +
    labs(x='Dose amount', y='Tooth length')
p2 <- ggplot(data=ToothGrowth, mapping=aes(factor(dose), len, colour=supp)) +
    geom_boxplot() +
    labs(x='Dose amount', y='Tooth length')
grid.arrange(p1, p2, ncol=2, top='Tooth length by supplement type and dose amount')
```

```{r tTestBoth, echo=F, results='asis', anchor='Table'}
dt <- ToothGrowth %>%
    mutate(group=paste(supp,dose,sep='-')) %>%
    select(len, group)
out <- with(dt, pairwise.t.test(x=len, g=group, p.adjust.method='bonferroni'))
outPValue <- out$p.value
print(xtable(outPValue, display=rep('E', ncol(outPValue) + 1),
             caption=paste('Result of multiple t-test of the comparison of',
                           'tooth growth by supplement type and dose amount.')))
```


## Conclusions

After these analysis we can conclude that dosing vitamin C through orange juice
is more effective than ascorbic acid and also that bigger doses affect tooth
growth more than smaller ones. In some cases the effect provided by a dose of
ascorbic acid can be achieved by a smaller dose of orange juice.

\clearpage

## Appendix

This appendix contains the code that has been used to generate the above
presented plots and tables. Here's the required libraries loaded:

```{r tablePlotslibrariesCode, ref.label='tablePlotslibraries', eval=F}
```

Code 1:
```{r bySupplementTypePlotCode, ref.label='bySupplementTypePlot', eval=F, anchor='Code'}
```

Code 2:
```{r byDoseAmountPlotCode, ref.label='byDoseAmountPlot', eval=F, anchor='Code'}
```

Code 3:
```{r byBothPlotCode, ref.label='byBothPlot', eval=F, anchor='Code'}
```

Code 4:
```{r datasetSummaryCode, ref.label='datasetSummary', eval=F, anchor='Code'}
```

Code 5:
```{r tTestOjGreaterVcCode, ref.label='tTestOjGreaterVc', eval=F, anchor='Code'}
```

Code 6:
```{r tTestBiggerDosesGreaterSmallerDosesCode, ref.label='tTestBiggerDosesGreaterSmallerDoses', eval=F, anchor='Code'}
```

Code 7:
```{r cohenTestBiggerDosesGreaterSmallerDosesCode, ref.label='cohenTestBiggerDosesGreaterSmallerDoses', eval=F, anchor='Code'}
```

Code 8:
```{r tTestBothCode, ref.label='tTestBoth', eval=F, anchor='Code'}
```
